---
title: "Anova"
output: html_document
---

title: "ANOVA for Journalists: Analyzing College Admissions Test Scores"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this exercise, we'll explore a common journalism question: *Are there significant differences in standardized test scores across different types of colleges?*

We'll use ANOVA (Analysis of Variance) to determine if the observed differences in SAT scores across three college types (Public, Private, and Community College) represent real patterns or just random variation.

## Task 1: Load Libraries and Create Sample Data

```{r}
library(tidyverse)
```

```{r create-data}
# Set seed for reproducibility
set.seed(123)

# Create simulated SAT scores for admitted students across three college types
# Scale is 400-1600 for total SAT
public_college <- rnorm(100, mean = 1350, sd = 120)
private_college <- rnorm(100, mean = 1450, sd = 100)
community_college <- rnorm(100, mean = 1050, sd = 150)

# Combine into a single dataframe
admissions_data <- data.frame(
  college_type = factor(rep(c("Public", "Private", "Community"), each = 100)),
  sat_score = c(public_college, private_college, community_college)
)

# Ensure SAT scores are within realistic range (400-1600)
admissions_data$sat_score <- pmin(pmax(admissions_data$sat_score, 400), 1600)

# Quick look at the data
head(admissions_data)
```

## Task 2: Calculate Basic Statistics

```{r summary-stats}
# Summary statistics by college type
admission_summary <- admissions_data |>
  group_by(college_type) |>
  summarize(
    mean_score = mean(sat_score),
    median_score = median(sat_score),
    sd_score = sd(sat_score),
    min_score = min(sat_score),
    max_score = max(sat_score)
  )

print(admission_summary)
```

**Task 3** Describe the data

Write a couple of sentences describing the differences in SAT scores between types of colleges, using the descriptive statistics we just generated.

The differences in the SAT scores between colleges shows that private schools have the highest SAT average, coming in at 1435. Private schools also have the lowest standard deviation, meaning that the variance in scores is smaller and more people scored similarly. The minimum score doe community college was the lowest out of all of the colleges, being 786. Community college overall had scores that were significantly lower than private and public.


## Task 4: Visualize the Data

Let's create visualizations to see the patterns:

```{r visualize-distributions}
# Create density plot showing overlapping distributions
ggplot(admissions_data, aes(x = sat_score, fill = college_type)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "SAT Score Distributions by College Type",
    subtitle = "Are these differences statistically significant?",
    x = "SAT Score (400-1600 scale)",
    y = "Density",
    fill = "College Type"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

```{r boxplot}
# Create boxplot for another view
ggplot(admissions_data, aes(x = college_type, y = sat_score, fill = college_type)) +
  geom_boxplot() +
  labs(
    title = "SAT Scores by College Type",
    x = "College Type",
    y = "SAT Score"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")
```

## Task 5: Perform ANOVA Test

```{r anova-test}
# Run one-way ANOVA
anova_result <- aov(sat_score ~ college_type, data = admissions_data)
summary(anova_result)
```

The F-value is huge (284.3). The F-value measures the ratio of between-group variance to within-group variance, and this high value suggests the differences between college types far outweigh the variation within each type. The p-value is very low, indicating the differences between college types are statistically significant at any reasonable significance level. The three asterisks (***) highlight that this is significant at the 0.001 level. So we can say that with confidence that there are real differences in SAT scores between the three types of colleges.

## Task 5: Compare the differences between specific pairs

```{r post-hoc}
# Tukey's HSD test to see which specific pairs of college types differ
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

The Tukey multiple comparisons test shows highly significant differences in SAT scores between all college types:

* Private vs. Community: Private colleges have SAT scores that are 366.96 points higher than community colleges on average (p < 0.0001)
* Public vs. Community: Public colleges have SAT scores that are 292.55 points higher than community colleges on average (p < 0.0001)
* Public vs. Private: Public colleges have SAT scores that are 74.40 points lower than private colleges on average (p = 0.000021)

All comparisons show extremely small p-values, indicating these differences are statistically significant at any conventional significance level. The 95% confidence intervals (shown in the "lwr" and "upr" columns) do not include zero for any comparison, further confirming these differences are statistically significant.

## Task 6: Report the Findings

Based on our analysis, we can now answer:

1. Is there a significant difference in SAT scores across college types?
There is a significant difference in SAT scores across college types. Private colleges have the highest scores, public colleges have slightly lower scores, and community colleges have the lowest scores. We know from the anova test that the differences are not a coincidence and are reflective of the schools.

2. Which specific college types differ from each other?

The biggest difference in scores is between the community college and private college. The private and public schools mean score is hundreds of points larger than the community college.

3. How large are these differences?

These differences are very large, especially between the private and community colleges as I mentioned. The max SAT score at community college is less than the average at the private college. This shows that the two colleges are on different sides of the score spectrum. 


Answer those questions, providing evidence for your answers, and then write a headline and a lead paragraph reporting these findings using active voice and focusing on the differences across types.

Maryland private college has a nearly 400 point average difference in SAT scores, compared to community college.

This difference is reflective of the overall accepted college population. The maximum SAT score at community college is still less than the average at the private college.

