---
title: "midterm_study_guide"
output: html_document
---

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```


### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  mutate(
```


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(june_change = (Jun - average_incidents) / average_incidents * 100) |> 
  mutate(july_change = (Jul - average_incidents) / average_incidents * 100) |> 
  mutate(august_change = (Aug - average_incidents) / average_incidents * 100)
```





### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(salary),
            median_salary = median(salary)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
  summarize(mean_salary = mean(salary),
            median_salary = median(salary),
            min_salary = min(salary),
            max_salary = max(salary)
)
```



### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> 
mutate(violent_rate_2019 = (violent2019/pop2019) * 10000,
  property_rate_2019 = (property2019/pop2019) * 10000,
  violent_rate_2020 = (violent2020/pop2020) * 10000,
  property_rate_2020 = (property2020/pop2020) * 10000)
```



### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>
  mutate(property_change = ((property_rate_2020 - property_rate_2019)/property_rate_2020) * 100,
  violent_change = ((violent_rate_2020 - violent_rate_2019)/violent_rate_2020) * 100)
  

```


### Task 10: Percentiles
```{r}
md_cities_percentiles <- md_cities_change |> 
  mutate(violent_percentile = ntile (violent_rate_2020, 100))

md_cities_percentiles |> 
  summarise(IQR = IQR(violent_rate_2020))
```



### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarise(mean = mean(danceability), sd = sd(danceability))
```


### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> 
  ggplot() +
  geom_histogram(aes(x = danceability))
```



### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> 
  ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color="red" , linetype = "dashed", size = 1) 
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs |> 
  ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color="red" , linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color= "blue" , linetype = "dashed", size = 1)+
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color= "blue" , linetype = "dashed", size = 1)
```




### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}

sample5 <- mm_data |> 
  sample_n(5)

sample10 <- mm_data |> 
  sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 |> summarize(mean=mean(black), sd=sd(black))
sample10 |> summarize(mean=mean(black), sd=sd(black))
```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
 mm_data |> summarize(mean=mean(black), sd=sd(black)) 
```



### Task 3: Calculate the correlation coefficient

Use the `cor()` function to calculate the correlation coefficient between the DJIA and GDP.

```{r}
gdp_djia |> 
  summarize(correlation = cor(DJIA, GDP))
```


### Task 4: Create a scatterplot

Use the `ggplot()` function to create a scatterplot of the DJIA and GDP.

```{r}
gdp_djia |> 
  ggplot() +
  geom_point(aes(x = DJIA, y = GDP)) +
  geom_smooth(aes(x = DJIA, y = GDP), method = 'lm') +
  geom_text(aes(x = DJIA, y = GDP, label = Quarter)) +
  labs(title = "DJIA vs. GDP",subtitle = "What's the relationship?",
       x = "DJIA",
       y = "GDP")
```

