---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mm_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}

sample5 <- mm_data |> 
  sample_n(5)

sample10 <- mm_data |> 
  sample_n(10)
```

### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 |> summarize(mean=mean(black), sd=sd(black))
sample10 |> summarize(mean=mean(black), sd=sd(black))
```

### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
 mm_data |> summarize(mean=mean(black), sd=sd(black)) 
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The means are all relatively similar, they were all between 73-77, however the standard deviations were all over the place at 
15, 26, and 19.
### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

From what I found in the data, you cannot rely on random sampling to get you an authentic perspective on your study. The mean and standard deviations were all too different. If I were to write a news story just based off of the 5 or 10 sample size then it would be fake news. If you want to get the best results you should not go random for your sample size.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.