---
title: "Final_Project"
output: html_document
---

```{r}
library(tidyverse)
```

```{r}
college_statistics <- read.csv("https://raw.githubusercontent.com/rachelbrock/jour405_spring/refs/heads/main/project/Final%20Project%20Data%20-%20Sheet1%20(1).csv")
```
Disclaimer: For some reason read_csv was NOT working and read.csv DID work to load my data. I am not sure why that is.

Final Project Goal:

The aim of my final project is to study the relationship between acceptance rate and graduation rate in the Big10 conference. Through R, I can find the average rates of acceptance and graduation for each school. I can also analyze if lower graduation rates mean that a school has a higher graduation rate can conclude if these trends are just a coincidence or if there is enough data to prove statistical significance (ANOVA test). I would also be open to doing more tests, but I am not one hundred percent sure what else I could do. The columns that will be most important will be the ones along the top row that show the acceptance and graduation rates for each year.  

Null Hypothesis: Schools with low graduation rates do not have higher graduation rates than those with higher acceptance rates.

Alternative Hypothesis: Schools with low graduation rates do have higher graduation rates than those with higher acceptance rates.

Data Description:

I found my data through a variety of sources which included the universities' research offices, college evaluation websites (ex.niche) and Google. It was impossible to get all of my data from a singular source, so I complied what I found from each source on a Google Sheets document. Because there is both a 4 year graduation rate and 6 year graduation rate for colleges, I chose to focus my analysis on the 4 year graduation rate. The time periods covered are 2020, 2022, and 2024 to give some variety to the data. I believe that the pandemic overall helped to increase acceptance rates during 2020 and, in general, most acceptance rates have gone down since then. 
