---
title: "Final_Project"
output: html_document
---

```{r}
library(tidyverse)
library(ggrepel)
```

```{r}
college_statistics <- read_csv("https://raw.githubusercontent.com/rachelbrock/jour405_spring/refs/heads/main/project/Final%20Project%20Data%20-%20Sheet1%20(1).csv")

college_statistics_long <- college_statistics |>
  pivot_longer(
    cols = -College,
    names_to = "measure",
    values_to = "rate"
  ) |>
  # Separate the measure column to extract metric type and year
  separate(measure, into = c("metric", "year"), sep = "_Rate_") |>
  # Fix the metric names
  mutate(
    metric = case_when(
      metric == "Acceptance" ~ "acceptance_rate",
      metric == "4_Year_Graduation" ~ "graduation_rate"
    )
  ) |>
  # Convert to a wider format with separate columns for each metric
  pivot_wider(
    names_from = metric,
    values_from = rate
  )
```

Summary Stats (mean and standard deviation)

```{r}
college_statistics |> summarise(mean = mean(Acceptance_Rate_2020), sd = sd(Acceptance_Rate_2020))
college_statistics |> summarise(mean = mean(Acceptance_Rate_2022), sd = sd(Acceptance_Rate_2022))
college_statistics |> summarise(mean = mean(Acceptance_Rate_2024), sd = sd(Acceptance_Rate_2024))
```
```{r}
library(dplyr)

summary_stats <- college_statistics_long  |>
  group_by(year) |>
  summarize(
    MeanAcceptance = mean(acceptance_rate, na.rm = TRUE),
    SDAcceptance = sd(acceptance_rate, na.rm = TRUE),
    MeanGraduation = mean(graduation_rate, na.rm = TRUE),
    SDGraduation = sd(graduation_rate, na.rm = TRUE)
  )
print(summary_stats)
```


ANOVA TEST

```{r}
college_statistics_long$acceptance_group <- cut(
  college_statistics_long$acceptance_rate,
  breaks = 3,
  labels = c("Low", "Medium", "High")
)

anova_model <- aov(graduation_rate ~ acceptance_group, data = college_statistics_long)
summary(anova_model)


```


Do schools with low, medium, or high acceptance rates have significantly different graduation rates?

Grouping schools into 3 levels of acceptance rate (Low, Medium, High), and then checking if the average graduation rate is different across those 3 groups.

Null hypothesis (H₀): The mean graduation rates are equal across all acceptance rate groups.

Alternative hypothesis (H₁): At least one group has a different mean graduation rate.

If the p-value from summary(anova_model) is:

< 0.05 → You reject the null hypothesis. There's evidence that graduation rates differ by acceptance group.

≥ 0.05 → You do not reject the null. No significant difference found.


P value found: 0.0000672

This means: 

There is strong statistical evidence to reject the null hypothesis.

➤ Graduation rates are significantly different between at least one pair of acceptance rate groups (Low, Medium, High).



Scatter Plot: A scatter plot is a great way to visualize the relationship between acceptance rate and graduation rate.

```{r}
ggplot(college_statistics_long, aes(x = acceptance_rate, y = graduation_rate)) +
  geom_point(color = "steelblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1) +
  geom_text_repel(aes(label = College), size = 3) +
  labs(
    title = "Acceptance Rate vs Graduation Rate (Labeled with Line of Best Fit)",
    x = "Acceptance Rate",
    y = "Graduation Rate"
  ) +
  theme_minimal()
```



Linear Regression:


```{r}
model <- lm(graduation_rate ~ acceptance_rate, data = college_statistics_long)
summary(model)
```



lm() fits the linear regression model.

graduation_rate ~ acceptance_rate means you're predicting graduation rate based on acceptance rate.

summary(model) gives you:

Coefficients (slope and intercept)

R-squared (how well the model fits)

p-value (to see if the relationship is statistically significant)



```{r}
library(ggplot2)
library(ggrepel)

ggplot(college_statistics_long, aes(x = acceptance_rate, y = graduation_rate)) +
  geom_point(color = "steelblue", size = 3) +  # data points
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 1) +  # regression line with confidence interval
  labs(
    title = "Linear Regression: Graduation Rate vs Acceptance Rate",
    x = "Acceptance Rate",
    y = "Graduation Rate"
  ) +
  theme_minimal()
```


Dots = Each school

Red line = Best-fit line from linear regression

Shaded area = Confidence interval (can remove with se = FALSE)

Labels = School names (won’t overlap thanks to ggrepel)


Link to Chat GPT conversation: https://chatgpt.com/share/68168fde-6eb8-8005-99db-5f616ed79cb5 






 





